### 2011-10-31 docs/src/currydoc.tex
CurryDoc improvements

\documentclass[11pt,fleqn]{article}

\usepackage{latexsym}
\usepackage{makeidx}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

%%% ------------------------------------------------------------------

\usepackage[colorlinks,linkcolor=blue]{hyperref}
\hypersetup{bookmarksopen=true}
\hypersetup{bookmarksopenlevel=0}
\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
\hypersetup{pdfauthor={Michael Hanus}}
%\hypersetup{pdfstartview=Title}
\hypersetup{pdfstartview=FitH}
\usepackage{thumbpdf}

%%% ------------------------------------------------------------------

\setlength{\textwidth}{16.5cm}
\setlength{\textheight}{23cm}
\renewcommand{\baselinestretch}{1.1}
\setlength{\topmargin}{-1cm}
\setlength{\oddsidemargin}{0cm}
\setlength{\evensidemargin}{0cm}
\setlength{\marginparwidth}{0.0cm}
\setlength{\marginparsep}{0.0cm}

\newlength{\figurewidth}
\setlength{\figurewidth}{\textwidth}
\addtolength{\figurewidth}{-0.4cm}

% font for program texts
\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
\newcommand{\codefont}{\tt}

% environment for typing program texts:
\makeatletter
\newenvironment{prog}{\par\vspace{0.7ex}
\setlength{\parindent}{1.0cm}
\setlength{\parskip}{-0.1ex}
\obeylines\@vobeyspaces\tt}{\vspace{0.7ex}\noindent
}
\makeatother
\newcommand{\startprog}{\begin{prog}}
\newcommand{\stopprog}{\end{prog}\noindent}

% program text in normal text
\newcommand{\code}[1]{\mbox{\codefont #1}}

% program text in normal text with apostrophs
\newcommand{\ccode}[1]{``\mbox{\codefont #1}''}

\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index

\newcommand{\labs}{\mbox{\tt\char92}}  % lambda abstraction in Curry
\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
\newcommand{\cb}{CurryBrowser\xspace}

% allow underscores in programs:
\catcode`\_=\active
\let_=\sb
\catcode`\_=12

% produce an index:
\makeindex

\begin{document}
\sloppy

\begin{titlepage}
\pdfbookmark[1]{Title}{Title}
\begin{center}
\fbox{
\begin{minipage}[t]{\figurewidth}
\begin{center}\vspace{10ex}
{\Huge\bf KiCS2 0.1}\\[4ex]
{\huge The Kiel Curry System (Version 2)}\\[7ex]
{\huge User Manual}\\[4ex]
Version of October 28, 2011\\[6ex]
\Large
Michael Hanus$^1$ [editor] \\[3ex]
{\large Additional Contributors:}\\[2ex]
Bernd Bra\ss{}el$^2$ \\
Bj\"orn Peem\"oller$^3$\\
Fabian Reck$^4$ \\[4ex]
\normalsize
(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
\end{center}
\end{minipage}}
\end{center}
\end{titlepage}

\pdfbookmark[1]{Contents}{Contents}
\tableofcontents

\newpage

\addcontentsline{toc}{section}{Preface}
\section*{Preface}

This document describes KiCS2 (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
Version \textbf{2}),
an implementation of the multi-paradigm language Curry
\cite{Hanus97POPL,Hanus06Curry} that is
based on compiling Curry programs into Haskell programs.
Curry is a universal programming language aiming at the amalgamation
of the most important declarative programming paradigms,
namely functional programming and logic programming.  
Curry combines in a seamless way features from functional programming
(nested expressions, lazy evaluation, higher-order functions),
logic programming (logical variables, partial data structures,
built-in search), and concurrent programming (concurrent evaluation
of constraints with synchronization on logical variables).
The current KiCS2 implementation does not support concurrent
constraints. Alternatively, one can write distributed applications
by the use of sockets that can be registered and accessed
with symbolic names.
Moreover, KiCS2 also supports
the high-level implementation of
graphical user interfaces and web services
(as described in more detail in
\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).

We assume familiarity with the ideas and features
of Curry as described in the Curry language definition \cite{Hanus06Curry}.
Therefore, this document only explains the use of the different
components of KiCS2
and the differences and restrictions of KiCS2
(see Section~\ref{sec-restrictions})
compared with the language Curry (Version 0.8.2).
The basic ideas of the implementation of KiCS2
can be found in
\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.

\bigskip

\subsection*{Acknowledgements}

This work has been supported in part by the
DFG grants Ha 2457/5-1 and Ha 2457/5-2.

\newpage

\section{Overview of KiCS2}

\subsection{General Use}
\label{sec-general}

This version of KiCS2 has been developed and
tested on Linux systems.
In principle, it should be also executable on other
platforms on which a Haskell implementation (Glasgow Haskell Compiler
and Cabal) exists, like in many Linux distributions, Sun Solaris,
or Mac OS X systems
(see the file \code{INSTALL.html} in the KiCS2 directory
for a description of the necessary software to install KiCS2).

All executables required to use the different components
of KiCS2 are stored in the directory \code{$kics2home$/bin}
(where $kics2home$ is the installation directory of the complete
KiCS2 installation). You should add this directory
to your path (e.g., by the \code{bash} command
\ccode{export PATH=$kics2home$/bin:\$PATH}).

The source code of the Curry program
must be stored in a file with the suffix \ccode{.curry},
e.g., \code{prog.curry}. 
Literate programs must be stored in files with the extension \ccode{.lcurry}.

Since the translation of Curry programs with KiCS2 creates
some auxiliary files (see Section~\ref{sec-auxfiles} for details),
you need write permission
in the directory where you have stored your Curry programs.
Moreover, the current implementation also recompiles
system libraries according to the setting of some options.
Therefore, the KiCS2 system should be locally installed
in your user account.
The auxiliary files for all Curry programs in the current
directory can be deleted by the command\pindex{cleancurry}
\startprog
cleancurry
\stopprog
(this is a shell script stored in the \code{bin} directory of the
KiCS2 installation, see above).
The command
\startprog
cleancurry -r
\stopprog
also deletes the auxiliary files in all subdirectories.

\subsection{Restrictions on Curry Programs}
\label{sec-restrictions}

There are a few minor restrictions on Curry programs
when they are processed with KiCS2:
\begin{itemize}
\item \index{singleton variables}\index{variables!singleton}
\emph{Singleton variables}, i.e., variables that occur only once
in a rule, should be denoted as an anonymous variable \ccode{_},
otherwise the parser will print a warning since this is a
typical source of programming errors.
\item
KiCS2 translates all \emph{local declarations} into global functions with
additional arguments (``lambda lifting'', see Appendix~D of the
Curry language report).
Thus, in the various run-time systems, the definition of
functions with local declarations look different from
their original definition (in order to see the result
of this transformation, you can use the \cb, see
Section~\ref{sec-currybrowser}).
\item \index{tabulator stops}
Tabulator stops instead of blank spaces in source files are
interpreted as stops at columns 9, 17, 25, 33, and so on.
%\item Threads created by a concurrent conjunction are not executed
%in a fair manner (usually, threads corresponding to leftmost constraints
%are executed with higher priority).
\item
Encapsulated search\index{encapsulated search}: In order
to allow the integration of non-deterministic computations
in programs performing I/O at the top-level, KiCS2 provides
the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
and \code{SearchTree} (Section~\ref{Library:SearchTree})
where one can defined search operators and various search strategies.
The general definition of encapsulated search of the Curry report
\cite{HanusSteiner98PLILP} is not supported.
\item
There is currently no general connection to external constraint solvers.
\end{itemize}

\subsection{Modules in KiCS2}
\label{sec-modules}

The current implementation of KiCS2 supports only flat module names,
i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
In order to allow the structuring of modules in different directories,
KiCS2 searches for imported modules in various directories.
By default, imported modules are searched in the directory
of the main program and the system module directories
\ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
This search path can be extended
by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
(which can be also set in a KiCS2 session by the option
\ccode{:set path}\pindex{path}\pindex{:set path},
see below)
to a list of directory names separated by colons (\ccode{:}).
In addition, a local standard search path
can be defined in the \ccode{.kics2rc} file
(see Section~\ref{sec-customization}).
Thus, modules to be loaded are searched in the following
directories (in this order, i.e., the first occurrence of a module file
in this search path is imported):
\begin{enumerate}
\item Current working directory (\ccode{.}) or directory prefix
of the main module (e.g., directory \ccode{/home/joe/curryprogs}
if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
\item The directories enumerated in the environment variable \code{CURRYPATH}.
\item The directories enumerated in the \ccode{.kics2rc} variable
      \ccode{libraries}.
\item The directories \ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
\end{enumerate}
Note that the standard prelude (\code{$kics2home$/lib/Prelude.curry})
will be always implicitly imported to all modules if a module
does not contain an explicit import declaration for the module
\code{Prelude}.

\newpage

\section{Using the Interactive Environment of KiCS2}
\label{sec-kics2}

This section describes the interactive environment
KiCS2\index{KiCS2}
that supports the development of applications written in Curry.
The implementation of KiCS2 contains also a separate compiler
which is automatically invoked by the interactive environment.

\subsection{Invoking KiCS2}
\label{sec:invoke-kics2}

To start KiCS2, execute the command
\ccode{kics2}\pindex{kics2}
(this is a shell script stored in
\code{$kics2home$/bin} where $kics2home$ is the installation directory
of KiCS2).
When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
the prelude (\code{$kics2home$/lib/Prelude.curry})
is already loaded, i.e., all definitions in the prelude are accessible.
Now you can type various commands (see next section)
or an expression to be evaluated.

One can also invoke KiCS2 with parameters.
These parameters are usual a sequence of commands
(see next section) that are executed before the user interaction
starts. For instance, the invocation
\startprog
kics2 :load Mod :add List
\stopprog
starts KiCS2, loads the main module \code{Mod}, and adds the additional
module \code{List}.
The invocation
\startprog
kics2 :load Mod :eval config
\stopprog
starts KiCS2, loads the main module \code{Mod}, and evaluates
the operation \code{config} before the user interaction starts.
As a final example, the invocation
\startprog
kics2 :load Mod :save :quit
\stopprog
starts KiCS2, loads the main module \code{Mod}, creates an executable,
and terminates KiCS2. This invocation could be useful in ``make'' files
for systems implemented in Curry.

\subsection{Command of KiCS2}
\label{sec:kics2-commands}

The {\bf most important commands} of KiCS2 are
(it is sufficient to type a unique prefix of a command if it is unique,
e.g., one can type \ccode{:r} instead of \ccode{:reload}):

\begin{description}
\item[\fbox{\code{:help}}]\pindex{:help}
Show a list of all available commands.

\item[\fbox{\code{:load $prog$}}]\pindex{:load}
Compile and load the program stored in \code{$prog$.curry}
together with all its imported modules.
% If this file does not exist, the system looks for a FlatCurry
% file \code{$prog$.fcy} and compiles from this intermediate representation.
% If the file \code{$prog$.fcy} does not exists, too, the system looks
% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
% XML representation (compare command \ccode{:xml}\pindex{:xml}),
% translates this into a FlatCurry file \code{$prog$.fcy}
% and compiles from this intermediate representation.

\item[\fbox{\code{:reload}}]\pindex{:reload}
Recompile all currently loaded modules.

\item[\fbox{\code{:add} $m$}]\pindex{:add}
Add module $m$ to the set of currently loaded modules
so that its exported entities are available in the top-level environment.

\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
and show the computed results.
In the default mode, all results of
non-deterministic computations are printed.
One can also print first one result and the next result
only if the user requests it. This behavior can be set by
the option \code{interactive}\index{interactive} (see below).

\textbf{Free variables in initial expressions} must be declared as
in Curry programs, i.e., in order to see the results of their bindings,
they must be introduced by a \ccode{where\ldots{}free} declaration.
For instance, one can write
\startprog
xs++ys\,=:=\,[1,2]  where xs,ys free
\stopprog
in order to obtain the following three possible bindings:
\startprog
\{xs = [], ys = [1,2]\} Success
\{xs = [1], ys = [2]\} Success
\{xs = [1,2], ys = []\} Success
\stopprog
Without these declarations, an error is reported in order to
avoid the unintended introduction of free variables in initial expressions
by typos.

If the free variables in the initial goal are of a polymorphic type,
as in the expression
\startprog
xs++ys=:=[z] where xs,ys,z free
\stopprog
they are specialized to the type \ccode{()}
(since the current implementation of KiCS2 cannot compute
with polymorphic logic variables).

\item[\fbox{:eval $expr$}]\pindex{:eval}
Same as $expr$. This command might be useful when putting
commands as arguments when invoking \code{kics2}.

% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
% Define the identifier $x$ as an abbreviation for the expression $expr$
% which can be used in subsequent expressions. The identifier $x$
% is visible until the next \code{load} or \code{reload} command.

\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
\end{description}
%
\bigskip
%
There are also a number of {\bf further commands} that are often
useful:
%
\begin{description}
\item[\fbox{\code{:type $expr$}}]\pindex{:type}
Show the type of the expression $expr$.

\item[\fbox{\code{:programs}}]\pindex{:programs}
Show the list of all Curry programs that are available in the load path.

\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
Change the current working directory to $dir$.

\item[\fbox{\code{:edit}}]\pindex{:edit}
Load the source code of the current main module into a text editor.
If the variable \code{editcommand} is set in the
configuration file \ccode{.kics2rc}
(see Section~\ref{sec-customization}),
its value is used as an editor command, otherwise
the environment variable \ccode{EDITOR} is used as the editor program.

\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
Load file $file$ into a text editor which is defined
as in the command \ccode{:edit}.

\item[\fbox{\code{:show}}]\pindex{:show}
Show the source text of the currently loaded Curry program.
If the variable \code{showcommand} is set in the
configuration file \ccode{.kics2rc}
(see Section~\ref{sec-customization}),
its value is used as a command to show the source text,
otherwise the \ccode{cat} is used.

\item[\fbox{\code{:show $m$}}]\pindex{:show}
Show the source text of module $m$ which must be accessible
via the current load path.

\item[\fbox{\code{:source $f$}}]\pindex{:source}
Show the source code of function $f$ (which must be visible
in the currently loaded module) in a separate window.

\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
Show the source code of function $f$ defined in module $m$
in a separate window.

\item[\fbox{\code{:browse}}]\pindex{:browse}
Start the CurryBrowser to analyze the currently loaded
module together with all its imported modules
(see Section~\ref{sec-currybrowser} for more details).

\item[\fbox{\code{:interface}}]\pindex{:interface}
Show the interface of the currently loaded
module, i.e., show the names of all imported modules,
the fixity declarations of all exported operators,
the exported datatypes declarations and the types
of all exported functions.

\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
Similar to \ccode{:interface}
but shows the interface of the module $m$
which must be in the load path of KiCS2.

\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
Show all calls to imported functions in the currently loaded module.
This might be useful to see which import declarations are really necessary.

\item[\fbox{\code{:set $option$}}]\pindex{:set}
Set or turn on/off a specific option
of the KiCS2 environment (see \ref{sec:options} for a description
of all options). Options are turned on by the prefix
\ccode{+} and off by the prefix \ccode{-}. Options that can only
be set (e.g., \code{path}) must not contain a prefix.

\item[\fbox{\code{:set}}]\pindex{:set}
Show a help text on the possible options
together with the current values of all options.

\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
program as an executable evaluating the main expression \ccode{main}.
The executable is stored in the file \code{Mod}
if \code{Mod} is the name of the currently loaded main module.

\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
but the expression $expr$ (typically: a call to the main
function) will be evaluated by the executable.

\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
The expression $expr$, which is typically of type \ccode{IO ()},
is evaluated in an independent process which runs in
parallel to the current KiCS2 process.
All output and error messages from this new process are suppressed.
This command is useful to test distributed Curry programs
where one can start
a new server process by this command. The new process
will be terminated when the evaluation of the expression $expr$
is finished.

\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.

\end{description}

\subsection{Option of KiCS2}
\label{sec:options}

The following options (which can be set by the command \ccode{:set})
are currently supported:

\begin{description}
\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
for loading modules to $path$.
Note that this search path is only used for loading modules
inside this invocation of KiCS2.

\item[\fbox{\code{prdfs }}]\pindex{prdfs}
Set the search mode to evaluate non-deterministic expressions
to primitive depth-first search (which is usually the fastest method
to print all non-deterministic values).

\item[\fbox{\code{dfs}}]\pindex{dfs}
Set the search mode to evaluate non-deterministic expressions
to depth-first search. Usually, all non-deterministic values
are printed with a depth-first strategy, but one can also
print only the first value or all values by interactively requesting them
(see below for these options).

\item[\fbox{\code{bfs}}]\pindex{bfs}
Similarly to \code{dfs} but use a breadth-first search strategy
to compute and print the values of the given expression.

\item[\fbox{\code{ids}}]\pindex{ids}
Similarly to \code{dfs} but use an iterative-deepening strategy
to compute and print the values of the initial expression.
The initial depth bound is 100 and the depth-bound is
doubled after each iteration.

\item[\fbox{\code{ids $n$}}]\pindex{ids}
Similarly to \code{ids} but use an initial depth bound of $n$.

\item[\fbox{\code{par}}]\pindex{par}
Similarly to \code{dfs} but use a parallel search strategy
to compute and print the values of the initial expression.
The system chooses an appropriate number
of threads according the current number of available processors.

\item[\fbox{\code{par $n$}}]\pindex{par}
Similarly to \code{par} but use $n$ parallel threads.

\item[\fbox{\code{choices}}]\pindex{choices}
Show the internal choice structure
(according to the implementation described in
\cite{BrasselHanusPeemoellerReck11})
resulting from the complete evaluation of the main expression.
This mode is only useful for debugging purposes.

\item[\fbox{\code{supply $i$}}]\pindex{supply}
Use the implementation $i$ for the identifier supply for choice structures
(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
Currently, the values \code{ghc} (default), \code{integer},
or \code{ioref} are supported for $i$.

\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
Set the verbosity level to $n$. The following values are allowed
for $n$:
\begin{description}
\item[$n=0$:] Do not show any messages (except for errors).
\item[$n=1$:] Show only messages of the front-end, like loading
of modules.
\item[$n=2$:]
Show also messages of the back end, like compilation messages
from the Haskell compiler.
\item[$n=3$:]
Show also intermediate messages and commands of the compilation
process.
\item[$n=4$:]
Show also all intermediate results of the compilation process.
\end{description}

\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
Turn on/off the interactive mode.
In the interactive mode, the next non-deterministic value
is only computed when the user requests it.
In this mode, the computation is performed in a separate
window created by the command \code{interactivecommand}
whose value is defined in \ccode{.kics2rc}
(see Section~\ref{sec-customization}).

\item[\fbox{\code{+/-first}}]\pindex{first}
Turn on/off the first-only mode.
In the first-only mode, only the first value
of the main expression is printed (instead of all values).

\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
Turn on/off the optimization of the target program.

\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
Turn on/off the binding mode.
If the binding mode is on (default),
then the bindings of the free variables of the initial expression
are printed together with the result of the expression.

\item[\fbox{\code{+/-time}}]\pindex{time}
Turn on/off the time mode. If the time mode is on,
the cpu time and the elapsed time
of the computation is always printed together with the result
of an evaluation.

\item[\fbox{\code{+/-ghci}}]
Turn on/off the ghci mode.
In the ghci mode, the initial goal is send to the interactive version
of the Glasgow Haskell Compiler. This might result in a slower
execution but in a faster startup time since the linker
to create the main executable is not used.

\item[\fbox{\code{copts $o$}}]\pindex{copts}
Define additional options passed to the KiCS2 compiler.
For instance, setting the option
\startprog
:set copts -O 0
\stopprog
has the effect that all optimizations performed by the KiCS2 compiler
are turned off.

\item[\fbox{\code{rts $o$}}]\pindex{rts}
Define additional run-time options passed to the executable
generated by the Glasgow Haskell Compiler, i.e., the parameters
\ccode{+RTS $o$ -RTS} are passed to the executable.
For instance, setting the option
\startprog
:set rts -H512m
\stopprog
has the effect that the minimum heap size is set to 512 megabytes.

\end{description}

\subsection{Source-File Options}

If the evaluation of operations in some main module loaded into
KiCS2 requires specific options, like an iterative-deepening
search strategy, one can also put these options into the source
code of this module in order to avoid setting these options
every time when this module is loaded.
Such
{\bf source-file options}\index{source-file option}\index{option!in source file}
must occur before the module header, i.e., before the first declaration
(module header, imports, fixity declaration, defining rules, etc)
occurring in the module.
Each source file option must be in a line of the form
\startprog
\{-\# KiCS2_OPTION $opt$ \#-\}
\stopprog
where $opt$ is an option that can occur in a \ccode{:set} command
(compare Section~\ref{sec:options}).
Such a line in the source code (which is a comment according to
the syntax of Curry)
has the effect that this option is set by the KiCS2 command
\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
as a main module. For instance, if a module starts with the
lines
\startprog
\{-\# KiCS2_OPTION ids \#-\}
\{-\# KiCS2_OPTION +ghci \#-\}
\{-\# KiCS2_OPTION v2 \#-\}
module M where
$\ldots$
\stopprog
then the load command \ccode{:load M} will also
set the options for iterative deepening, using \code{ghci}
and verbosity level 2.

\subsection{Command Line Editing}

In order to have support for line editing or history functionality
in the command line of KiCS2 (as often supported by the \code{readline}
library), you should have the Unix command \code{rlwrap} installed
on your local machine.
If \code{rlwrap} is installed, it is used by KiCS2 if called on a terminal.
If it should not be used (e.g., because it is executed
in an editor with \code{readline} functionality), one can
call KiCS2 with the parameter \ccode{--noreadline}
(which must occur as the first parameter).

\subsection{Customization}
\label{sec-customization}

In order to customize the behavior of KiCS2 to your own preferences,
there is a configuration file which is read by KiCS2 when it is invoked.
When you start KiCS2 for the first time, a standard version of
this configuration file is copied with the name
\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
into your home directory. The file contains definitions
of various settings, e.g., about showing warnings, using Curry extensions,
programs etc.
After you have started KiCS2 for the first time, look into this file
and adapt it to your own preferences.

\subsection{Emacs Interface}

Emacs is a powerful programmable editor suitable for program development.
It is freely available for many platforms
(see \url{http://www.emacs.org}).
The distribution of KiCS2 contains also a special
\emph{Curry mode}\index{Curry mode}\index{Emacs}
that supports the development of Curry programs in
the Emacs environment.
This mode includes support for syntax highlighting,
finding declarations in the current buffer, and
loading Curry programs into KiCS2
in an Emacs shell.

The Curry mode has been adapted from a similar mode for Haskell programs.
Its installation is described in the file \code{README}
in directory \ccode{$kics2home$/tools/emacs} which also contains
the sources of the Curry mode and a short description about
the use of this mode.

\newpage

\section{Extensions}
\label{sec-extensions}

KiCS2 supports some extensions in Curry programs that are not (yet)
part of the definition of Curry. These extensions are described below.

\subsection{Recursive Variable Bindings}

Local variable declarations (introduced by \code{let}\pindex{let}
or \code{where}\pindex{where}) can be (mutually) recursive in KiCS2.
For instance, the declaration
\startprog
ones5 = let ones = 1 : ones
         in take 5 ones
\stopprog
introduces the local variable \code{ones} which is bound
to a \emph{cyclic structure}\index{cyclic structure}
representing an infinite list of \code{1}'s.
Similarly, the definition
\startprog
onetwo n = take n one2
 where
   one2 = 1 : two1
   two1 = 2 : one2
\stopprog
introduces a local variables \code{one2} that represents
an infinite list of alternating \code{1}'s and \code{2}'s
so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.

\subsection{Functional Patterns}

Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
to code operations in a more readable way. Furthermore,
defining operations with functional patterns avoids problems
caused by strict equality (\ccode{=:=}) and leads to programs
that are potentially more efficient.

Consider the definition of an operation to compute the last element
of a list \code{xs} based on the prelude operation \ccode{++}
for list concatenation:
\startprog
last xs | ys++[y] =:= xs  = y   where y,ys free
\stopprog
Since the equality constraint \ccode{=:=} evaluates both sides
to a constructor term, all elements of the list \code{xs} are
fully evaluated in order to satisfy the constraint.

Functional patterns can help to improve this computational behavior.
A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
is a function call at a pattern position. With functional patterns,
we can define the operation \code{last} as follows:
\startprog
last (_++[y]) = y
\stopprog
This definition is not only more compact but also avoids the complete
evaluation of the list elements: since a functional pattern is considered
as an abbreviation for the set of constructor terms obtained by all
evaluations of the functional pattern to normal form (see
\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
definition is conceptually equivalent to the set of rules
\startprog
last [y] = y
last [_,y] = y
last [_,_,y] = y
\ldots
\stopprog
which shows that the evaluation of the list elements is not demanded
by the functional pattern.

In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
is interpreted as a functional pattern if $f$ is not a visible constructor
but a defined function that is visible in the scope of the pattern.

\subsection {Records}
\label{records}

A record is a data structure for bundling several data of various types.
It consists of typed data fields where each field is associated with
a unique label. These labels can be used to construct, select or update
fields in a record.

Unlike labeled data fields in Haskell, records are 
not syntactic sugar but a real extension of the
language\footnote{The current version allows to transform records
  into abstract data types. Future extensions may not have
  this facility.}.
The basic concept is described in \cite{Leijen05} but the current
version does not yet provide all features mentioned there. 
The restrictions are explained in Section~\ref{sec-restrinrecs}.

\subsubsection{Record Type Declaration}
\label{sec-recordtypedecl}

It is necessary to declare a record type before a record
can be constructed or used. The declaration has the following form:
\startprog
type $R$ $\alpha_1$ \ldots $\alpha_n$ = \{ $l_1$ :: $\tau_1$, \ldots, $l_m$ :: $\tau_m$ \}
\stopprog
It introduces a new $n$-ary record type $R$ which represents a
record consisting of $m$ fields. Each field has a unique label $l_i$ 
representing a value of the type $\tau_i$. Labels
are identifiers which refer to the corresponding
fields. The following examples define some record types:
\startprog
type Person = \{name :: String, age :: Int\}
type Address = \{person :: Person, street :: String, city :: String\}
type Branch a b = \{left :: a, right :: b\}
\stopprog
It is possible to summarize different labels which have the same
type. For instance, the record \code{Address} can also be declared as follows:
\startprog
type Address = \{person :: Person, street,city :: String\}
\stopprog
The fields can occur in an arbitrary order. The example above
can also be written as
\startprog
type Address = \{street,city :: String, person :: Person\}
\stopprog
The record type can be used in every type expression to represent
the corresponding record, e.g.
\startprog
data BiTree = Node (Branch BiTree BiTree) | Leaf Int
\stopprog
\startprog
getName :: Person -> String
getName \ldots
\stopprog

Labels can only be used in the context of
records. They do not share the name space with 
functions/constructors/variables or type constructors/type variables. 
For instance it is possible to use 
the same identifier for a label and a function at the same time. Label
identifiers cannot be shadowed by other identifiers.

Like in type synonym declarations, recursive or mutually 
dependent record declarations are not allowed. Records can only
be declared at the top level. Further restrictions are described in
section \ref{sec-restrinrecs}.

\subsubsection{Record Construction}
\label{sec-recordconstr}

The record construction generates a record with initial values for
each data field. It has the following form:
\startprog
\{ $l_1$ = $v_1$, \ldots, $l_m$ = $v_m$ \}
\stopprog
It generates a record where each label $l_i$ refers to the
value $v_i$. The type of the record results from the record type
declaration where the labels $l_i$ are defined.
A mix of labels from different
record types is not allowed. All labels must be specified with 
exactly one assignment. Examples for record constructions are
\startprog
\{name = "Johnson", age = 30\}     -- generates a record of type 'Person'
\{left = True, right = 20\}        -- generates a record of type 'Branch'
\stopprog
Assignments to labels can occur in an arbitrary order. For instance a
record of type \code{Person} can also be generated as follows:
\startprog
\{age = 30, name = "Johnson"\}     -- generates a record of type 'Person'
\stopprog
Unlike labeled fields in record type declarations, 
record constructions can be used in expressions without any restrictions
(as well as all kinds of record expressions). For instance the following
expression is valid:
\startprog
\{person = \{name = "Smith", age = 20\},   -- generates a record of
 street = "Main Street",                -- type 'Address'
 city   = "Springfield"\}
\stopprog

\subsubsection{Field Selection}
\label{sec-fieldsel}

The field selection is used to extract data from records. 
It has the following form:
\startprog
$r$ -> $l$
\stopprog
It returns the value to which the label $l$ refers to from the
record expression $r$. The label must occur in the declaration of
the record type of $r$.
An example for a field selection is:
\startprog
pers -> name
\stopprog
This returns the value of the label \code{name} from the record \code{pers}
(which has the type \code{Person}).
Sequential application of field selections are also possible:
\startprog
(addr -> person) -> age
\stopprog
The value of the label \code{age} is extracted from a record which itself
is the value of the label \code{person} in the record \code{addr}
(which has the type \code{Address}). When a field selection is used in
expressions, it has to be parenthesized.

\subsubsection{Field Update}
\label{sec-fieldupd}

Records can be updated by reassigning a new value to a label:
\startprog
\{$l_1$ := $v_1$, \ldots, $l_k$ := $v_k$ | $r$\}
\stopprog
The label $l_i$ is associated with the new value $v_i$ which
replaces the current value in the record $r$.
The labels must occur in the declaration 
of the record type of $r$. In contrast to record constructions,
it is not necessary to specify all labels of a record. 
Assignments can occur in an arbitrary order. It is not allowed to 
specify more than one assignment for a label in a record update.
Examples for record updates are:
\startprog
\{name := "Scott", age := 25 | pers\}
\{person := \{name := "Scott", age := 25 | pers\} | addr\}
\stopprog
In these examples \code{pers} is a record of type \code{Person} and \code{addr}
is a record of type \code{Address}.

\subsubsection{Records in Pattern Matching}
\label{sec-recsinpm}

It is possible to apply pattern matching to records (e.g., in functions,
let expressions or case branches). Two kinds of record patterns
are available:
\startprog
\{$l_1$ = $p_1$, \ldots, $l_n$ = $p_n$\}
\{$l_1$ = $p_1$, \ldots, $l_k$ = $p_k$ | _\}
\stopprog
In both cases each label $l_i$ is specified with a pattern $p_i$. 
All labels must occur only once in the record pattern.
The first case is used to match the whole record. Thus, all labels
of the record must occur in the pattern. 
The second case is used to match only a part of
the record. Here it is not necessary to specify all labels.
This case is represented by a vertical bar followed by the underscore
(anonymous variable). It is
not allowed to use a pattern term instead of the underscore.

When trying to match a record against a record pattern, the 
patterns of the specified labels are matched against 
the corresponding values in the record expression. On success, all pattern
variables occurring in the patterns are replaced by their actual expression.
If none of the patterns matches, the computation fails.

Here are some examples of pattern matching with records:
\startprog
isSmith30 :: Person -> Bool
isSmith30 \{name = "Smith", age = 30\} = True
\stopprog
\startprog
startsWith :: Char -> Person -> Bool
startsWith c \{name = (d:_) | _\} = c == d
\stopprog
\startprog
getPerson :: Address -> Person
getPerson \{person = p | _\} = p
\stopprog
As shown in the last example, a field selection can also be obtained
by pattern matching.

\subsubsection{Export of Records}
\label{sec-exprecs}

Exporting record types and labels is very similar to exporting
data types and constructors. There are three ways 
to specify an export:
\begin{itemize}
\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
  exports the record $R$ without any of its labels.
\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
  exports the record $R$ together with all its labels.
\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
\end{itemize}
%
Note that imported labels cannot be overwritten in record declarations
of the importing module. It is also not possible to import equal labels
from different modules.

\subsubsection{Restrictions in the Usage of Records}
\label{sec-restrinrecs}

In contrast to the basic concept in \cite{Leijen05}, KiCS2/Curry provides a
simpler version of records. Some of the features described there are
currently not supported or even restricted.

\begin{itemize}
\item Labels must be unique within the whole scope of the program.
  In particular, it is not allowed to define the same label within
  different records, not even when they are imported from other
  modules. However, it is possible to use equal identifiers for other
  entities without restrictions, since labels have an independent 
  name space.
\item The record type representation with labeled fields can only be
  used as the right-hand-side of a record type declaration. It is
  not allowed to use it in any other type annotation.
\item Records are not extensible or reducible. The structure of a
  record is specified in its record declaration and cannot be
  modified at the runtime of the program.
\item Empty records are not allowed.
\item It is not allowed  to use a pattern term
  at the right side of the vertical bar in a record pattern
  except for the underscore (anonymous pattern variable).
\item Labels cannot be sequentially associated with multiple values
  (record fields do not behave like stacks).
\end{itemize}

>>>>

\section{CurryDoc: A Documentation Generator for Curry Programs}
